Imprecise Datapath Design: An Overclocking Approach

### 2013-06-21 TRETS2013/TRETS_overclocking.tex
TRETS modified introduction of CSA section

% Title portion
\title{Imprecise Datapath Design: An Overclocking Approach}
\author{KAN SHI, DAVID BOLAND\\
and GEORGE A. CONSTANTINIDES
\affil{Imperial College London}
}

@@

\section{Carry Select Adder}\label{Section_CSA}
\subsection{Introduction}
Since the delay of RCA is determined by the length of carry chain, alternative adder architectures, such as carry select adder (CSA), have been proposed to boost performance. In a CSA, the carry chain is divided into multiple overlapped stages, while each stage contains two RCAs and two multiplexers, as shown in Fig.~\ref{Fig_CSA_Structure} and Fig.~\ref{Fig_CSA_SingleStage}, respectively. For a given input, two additions are performed simultaneously within a single stage where the carry input is zero and one seperately. One of these two results is then selected according to the actual carry input. Although this structure brings performance benefits, it costs extra hardware resources compared to a standard RCA because the carry chain is duplicated. Furthermore, in FPGA technology, multiplexers are expensive to implement. Due to this reason, we explore the trade-offs between silicon area, accuracy and performance of RCA and CSA in this section.
\begin{figure}[htbp]
        \centering
        \subfigure[A CSA with $s$ stages.]{
            \includegraphics[width=0.6\textwidth]{./Figures/CSA_structure.pdf}
            \label{Fig_CSA_Structure}
        }%\vspace{-2ex}
        \subfigure[The $i^{th}$ stage in the CSA.]{
        \centering
            \includegraphics[width=2.2in]{./Figures/CSA_SingleStage.pdf}
            \label{Fig_CSA_SingleStage}
        }
    \caption{The structure diagram of CSA.}
\end{figure}

@@

\subsection{Accuracy Benefits and Area Overhead in CSA}
We first verify the models for the RCA and CSA in terms of the maximum word-length under the given operating frequencies. For the CSA, the ratio $\mu_{mux}/\mu_c$ can be computed experimentally. We perform post place-and-route simulations on the CSA with 2 stages using Xilinx Virtex-6 FPGA. The delay of the $i^{th}$ stage $d_i$ in~(\ref{Eq:CSA_SingleStageDelay}) is recorded with respect to different word-lengths. The total word-length of CSA can be predicted through~(\ref{Eq:CSA_Timing}). In addition, the maximum word-lengths of the 2-stage CSA and RCA are obtained experimentally by increasing $n_{CSA}$ and $n_{RCA}$ respectively until errors are observed at the output. The comparison between the modeled value and the empirical results is illustrated in Fig.~\ref{Fig_CSA2stage_Timing}. It can be seen that our models for both RCA and CSA match well with the experimental results.

\begin{figure}[htbp]
        \centering
        \subfigure[\scriptsize The modeled value and the experimental results of the maximum word-length of RCA and CSA, with respect to various frequency requirements.]{
            \includegraphics[width=3.5in]{./Figures/CSA2stage_Model_Timing.eps}
            \label{Fig_CSA2stage_Timing}
        }
        \subfigure[\scriptsize Hardware resource usage for an RCA and a CSA with 2 and 3 stages.]{
            \includegraphics[width=3.5in]{./Figures/CSA3stage_Area.eps}
            \label{Fig_CSA3stage_Area}
        }
    \caption{A comparison between RCA and CSA in terms of the maximum word-length of input signal and the area consumption.}
\end{figure}

@@

As an example slice through the design space, in Fig.~\ref{Fig_CSA_LUT25} we record the mean value of error at outputs with respect to different operating frequencies for both design scenarios when the number of the available LUTs is set to 25. In this graph, we have labeled the optimum adder structure. Note that in this and all the following experiments within this section, in order to apply our models the input data is randomly generated following the uniform distribution.
%
\begin{figure}[t]
  \centering
  \includegraphics[width=3.3in]{./Figures/Error_LUT25.eps}
  \caption{A comparison between two design scenarios when the number of available LUT is 25. The RCA and CSA are investigated in the conventional scenario, while the RCA is explored in the proposed new scenario. The results are obtained from post place-and-route simulations on Xilinx Virtex-6 FPGAs.}
  \vspace{-.5ex}
  \label{Fig_CSA_LUT25}
\end{figure}

@@

We then perform similar experiments with a variety of area constraints. The optimum design methods with respect to different operating frequencies and area consumptions are demonstrated in Fig.~\ref{Fig_CSA_Tradeoff}. From this figure, several observations can be made. Firstly, if the frequency requirement is not large whilst the available area is large enough to implement a CSA in full precision, it will be the optimum design. This is expected from our earlier analysis in Fig.~\ref{Fig_CSA2stage_Timing}. Secondly, the 2-stage CSA is better than the 3-stage CSA when frequency is low, as it consumes less area although both achieve the same error expectation. Thirdly, only part of the CSA can be implemented under a tighter area budget, whereas the RCA still keeps full precision. In this case, area becomes the dominant factor and precision is lost for the CSA, meaning the RCA with overclocking is the optimum design method across almost the whole frequency domain in this situation. Finally, for very stringent area constraints, the word-length of RCA is also limited. This results in truncation errors initially for all design scenarios. However, the RCA with overclocking can still be employed as the optimum design, because it loses less precision than the CSA under the same area constraint.

\begin{figure}[t]
  \centering
  \includegraphics[width=3.5in]{./Figures/Tradeoff.eps}
  \caption{A demonstration of the optimum design methodology which achieves the minimum error at outputs with respect to a variety of frequency and area constraints.}
  \label{Fig_CSA_Tradeoff}
\end{figure}

@@

In our experiments, MRE is set ranging from $0.001\%$ to $50\%$. For a certain value of MRE, the design with the maximum operating frequency is selected as the optimum design. Moreover, the smallest design is the optimum one if multiple structures operate at the same frequency, with a certain area requirement. Fig.~\ref{Fig_CSA_Tradeoff_Error} can thus be obtained based on these criteria.

\begin{figure}[t]
    \centering
    \includegraphics[width=3.7in]{./Figures/Tradeoff_Error.eps}
    \caption{A demonstration of the optimum design methodology which runs at the fastest frequency with respect to a variety of accuracy and area constraints.}
    \label{Fig_CSA_Tradeoff_Error}
\end{figure}

@@

In this CCM structure, let the two inputs of the RCA be denoted by $A_S$ and $A_O$ respectively, which are both two's complement numbers. $A_S$ denotes the ``shifted signal", with zeros padded after LSB, while $A_O$ denotes the ``original signal" with MSB sign extension. For an $n$-bit input signal, it should be noted that an $n$-bit RCA is sufficient for this operation, because no carry will be generated or propagated when adding with zeros, as shown in Fig.~\ref{CCM_fig}.

\begin{figure}[htbp]
  \vspace{-2ex}
  \centering
  \includegraphics[width=\textwidth]{./Figures/CCM_CarryChain.pdf}
  \vspace{-2ex}
  \caption{Four possible carry chain types in a constant coefficient multiplier with $n$-bit inputs. The notion $s$ denotes the shifted bits and $BP$ denotes the binary point.}
  \label{CCM_fig}
\end{figure}

@@

\subsection{Experimental Setup} \label{sub:experimental_setup}
We initially build up a test framework on an FPGA. The general architecture is depicted in Fig.~\ref{Test Framework}. The main body of the test framework consists of the circuit under test (CUT), the test frequency generator and the control logic, as shown in the dotted box in Fig.~\ref{Test Framework}. The I/Os of the CUT are registered by the launch registers (LRs) and the sample registers (SRs), which are all triggered by the test clock. Input test vectors are stored in the on-chip memory during initialization. The results are sampled using Xilinx ChipScope. Finally, we perform an offline comparison of the output of the original circuit at the rated frequency with the output of the overclocked as well as the truncated designs using the same input vectors.
%
\begin{figure}[htbp]
  \centering
  \vspace{-3ex}
  \includegraphics[width=3.5in]{./Figures/TestFramework.pdf}
  %\vspace{-4.5ex}
  \caption{Test framework, which is composed of a measurement architecture (the dotted box) on an FPGA and an off-line comparator using software. The error-free data are obtained by either pre-computation or initial run with low frequencies.}
  \label{Test Framework}
  %\vspace{-1ex}
\end{figure}

@@

For example, the predicted frequencies and the actual frequencies of a $5^{th}$~order FIR filter using different word-lengths are depicted in Fig.~\ref{Fig:TimingMargin}. The ``actual'' maximum frequencies are computed by increasing the operating frequency from the rated value until errors are observed at the output; the maximum operating frequency with correct output is recorded for the current word-length. As can be seen in Fig.~\ref{Fig:TimingMargin}, the circuit can operate without errors at a much higher frequency in practice than predicted according to our experiments. A maximum speed differential of $3.2\times$ is obtained when the input signal is 5-bit.
%
\begin{figure}[ht]
  \centering
  \includegraphics[width=3.2in]{./Figures/Exp/FIR/RatedFrequency3.eps}
  %\vspace{-1.5ex}
  \caption{The maximum operating frequencies for different input word-lengths of an FIR filter. The dotted line depicts the rated frequency reported by the timing analysis tool. The solid line is obtained through real FPGA tests using our platform.}
  \label{Fig:TimingMargin}
  %\vspace{-3.5ex}
\end{figure}

@@

\section{Results and Discussion}\label{Section_Experiments}
\subsection{Case study: FIR filter}
We first assess the accuracy of our proposed models of errors. We record the value of MRE based on (\ref{Eq:MRE}) in a $5^{th}$~order FIR with respect to different operating frequencies. The modeled values of both overclocking error and truncation error of the FIR filter are presented in Fig.~\ref{Fig:FIR} (dotted lines), as well as the actual measurements on the FPGA (solid lines) with two types of input data. The results demonstrate that our models match well with the practical results obtained using the uniform independent inputs.
% FIR
\begin{figure}[tbp]
  \centering
%  \vspace{-3ex}
  \includegraphics[width=3.5in]{./Figures/Exp/FIR/FIR_Error_New_TCAD.eps}
%  \vspace{-5ex}
  \caption{A demonstration of two design perspectives with a $5^{th}$~order FIR filter, which is implemented on Virtex-6 FPGA. The modeled values of both overclocking errors and truncation errors are presented as dotted lines. The actual FPGA measurements are depicted using solid lines. Two types of inputs are employed in the overclocking scenario: the uniformly distributed data and the real image data from Lena.}
  \label{Fig:FIR}
\end{figure}

@@

In addition, we record the probability distribution of different length of carry chains using an 8-bit RCA with both data types, as shown in Fig.~\ref{Fig:CarryDistribution}. As the input data is 8-bit, the longest possible carry chain is of length 9-bit. It can be observed that when using Lena data,  the probability is only higher for carry chain with a length of 3-bit. While in other situations, using uniform data leads to higher probability, especially for long carry chain length. This finding explicitly demonstrates that longer carry chains happens with an even smaller probability when utilizing real data.

\begin{figure}[tbp]
  \centering
  \includegraphics[width=3.5in]{./Figures/CC_length2.eps}
  \caption{Probability distribution of different length of carry chains in a 8-bit RCA. For the uniform data, two inputs of RCA are randomly sampled from a uniform distribution. For the image data Lena, one input of RCA uses the original data and the other uses the delayed original data for several clock cycles.}
  \label{Fig:CarryDistribution}
\end{figure}

>>>>

% Title portion
\title{Imprecise Datapath Design: An Overclocking Approach}
\author{KAN SHI, DAVID BOLAND\\
and GEORGE A. CONSTANTINIDES
\affil{Imperial College London}
}
%\vspace{-5ex}

@@

\section{Carry Select Adder}\label{Section_CSA}
\subsection{Introduction}
As seen in~(\ref{Eq:Comparison RCA}), a factor of 2 reduction in error expection can be achieved by using the proposed approach in RCA. However, implementing our approach costs extra silicon area in comparison to the conventional approach, because we keep the original word-length $k$ in the new scenario instead of truncating to only $b-1$ bits in the traditional scenario. Besides our approach, there are existing approaches that could be employed to trade silicon area for low latency. For instance, alternative adder architectures, such as carry select adder (CSA), have been proposed to boost performance. In a CSA, the carry chain is divided into multiple overlapped stages, while each stage contains two RCAs and two multiplexers, as shown in Fig.~\ref{Fig_CSA_Structure} and Fig.~\ref{Fig_CSA_SingleStage}, respectively. For a given input, two additions are performed simultaneously within a single stage where the carry input is zero and one seperately. One of these two results is then selected according to the actual carry input. Although this structure brings performance benefits, it costs extra hardware resources compared to a standard RCA because the carry chain is duplicated. Furthermore, in FPGA technology, multiplexers are expensive to implement. Due to this reason, we explore the trade-offs between silicon area, accuracy and performance of RCA and CSA in this section.
\vspace{-7ex}
\begin{figure}[htbp]
        \centering
        \subfigure[A CSA with $s$ stages.]{
            \includegraphics[width=0.6\textwidth]{./Figures/CSA_structure.pdf}
            \label{Fig_CSA_Structure}
        }%\vspace{-2ex}
        \subfigure[The $i^{th}$ stage in the CSA.]{
        \centering
            \includegraphics[width=2.2in]{./Figures/CSA_SingleStage.pdf}
            \label{Fig_CSA_SingleStage}
        }
    \caption{The structure diagram of CSA.}
\end{figure}
\vspace{-2ex}

@@

\subsection{Accuracy Benefits and Area Overhead in CSA}
We first verify the models for the RCA and CSA in terms of the maximum word-length under the given operating frequencies. For the CSA, the ratio $\mu_{mux}/\mu_c$ can be computed experimentally. We perform post place-and-route simulations on the CSA with 2 stages using Xilinx Virtex-6 FPGA. The delay of the $i^{th}$ stage $d_i$ in~(\ref{Eq:CSA_SingleStageDelay}) is recorded with respect to different word-lengths. The total word-length of CSA can be predicted through~(\ref{Eq:CSA_Timing}). In addition, the maximum word-lengths of the 2-stage CSA and RCA are obtained experimentally by increasing $n_{CSA}$ and $n_{RCA}$ respectively until errors are observed at the output. The comparison between the modeled value and the empirical results is illustrated in Fig.~\ref{Fig_CSA2stage_Timing}. It can be seen that our models for both RCA and CSA match well with the experimental results.
%
\begin{figure}[htbp]
        \centering
        \vspace{-3ex}
        \subfigure[\scriptsize The modeled value and the experimental results of the maximum word-length of RCA and CSA, with respect to various frequency requirements.]{
            \includegraphics[width=3.5in]{./Figures/CSA2stage_Model_Timing.eps}
            \label{Fig_CSA2stage_Timing}
        }
        \subfigure[\scriptsize Hardware resource usage for an RCA and a CSA with 2 and 3 stages.]{
            \includegraphics[width=3.5in]{./Figures/CSA3stage_Area.eps}
            \label{Fig_CSA3stage_Area}
        }
        \vspace{-1.5ex}
    \caption{A comparison between RCA and CSA in terms of the maximum word-length of input signal and the area consumption.}
\end{figure}

@@

As an example slice through the design space, in Fig.~\ref{Fig_CSA_LUT25} we record the mean value of error at outputs with respect to different operating frequencies for both design scenarios when the number of the available LUTs is set to 25. In this graph, we have labeled the optimum adder structure. Note that in this and all the following experiments within this section, in order to apply our models the input data is randomly generated following the uniform distribution.
%
\begin{figure}[t]
  \centering
  \vspace{-2ex}
  \includegraphics[width=3.3in]{./Figures/Error_LUT25.eps}
  \vspace{-2ex}
  \caption{A comparison between two design scenarios when the number of available LUT is 25. The RCA and CSA are investigated in the conventional scenario, while the RCA is explored in the proposed new scenario. The results are obtained from post place-and-route simulations on Xilinx Virtex-6 FPGAs.}
  \vspace{-.5ex}
  \label{Fig_CSA_LUT25}
\end{figure}

@@

We then perform similar experiments with a variety of area constraints. The optimum design methods with respect to different operating frequencies and area consumptions are demonstrated in Fig.~\ref{Fig_CSA_Tradeoff}. From this figure, several observations can be made. Firstly, if the frequency requirement is not large whilst the available area is large enough to implement a CSA in full precision, it will be the optimum design. This is expected from our earlier analysis in Fig.~\ref{Fig_CSA2stage_Timing}. Secondly, the 2-stage CSA is better than the 3-stage CSA when frequency is low, as it consumes less area although both achieve the same error expectation. Thirdly, only part of the CSA can be implemented under a tighter area budget, whereas the RCA still keeps full precision. In this case, area becomes the dominant factor and precision is lost for the CSA, meaning the RCA with overclocking is the optimum design method across almost the whole frequency domain in this situation. Finally, for very stringent area constraints, the word-length of RCA is also limited. This results in truncation errors initially for all design scenarios. However, the RCA with overclocking can still be employed as the optimum design, because it loses less precision than the CSA under the same area constraint.
%
\begin{figure}[t]
  \centering
  \vspace{-2ex}
  \includegraphics[width=3.5in]{./Figures/Tradeoff.eps}
  \vspace{-2ex}
  \caption{A demonstration of the optimum design methodology which achieves the minimum error at outputs with respect to a variety of frequency and area constraints.}
  \label{Fig_CSA_Tradeoff}
\end{figure}

@@

In our experiments, MRE is set ranging from $0.001\%$ to $50\%$. For a certain value of MRE, the design with the maximum operating frequency is selected as the optimum design. Moreover, the smallest design is the optimum one if multiple structures operate at the same frequency, with a certain area requirement. Fig.~\ref{Fig_CSA_Tradeoff_Error} can thus be obtained based on these criteria.
%
\begin{figure}[t]
    \centering
    \vspace{-2ex}
    \includegraphics[width=3.7in]{./Figures/Tradeoff_Error.eps}
    \vspace{-2ex}
    \caption{A demonstration of the optimum design methodology which runs at the fastest frequency with respect to a variety of accuracy and area constraints.}
    \label{Fig_CSA_Tradeoff_Error}
\end{figure}

@@

In this CCM structure, let the two inputs of the RCA be denoted by $A_S$ and $A_O$ respectively, which are both two's complement numbers. $A_S$ denotes the ``shifted signal", with zeros padded after LSB, while $A_O$ denotes the ``original signal" with MSB sign extension. For an $n$-bit input signal, it should be noted that an $n$-bit RCA is sufficient for this operation, because no carry will be generated or propagated when adding with zeros, as shown in Fig.~\ref{CCM_fig}.
%
\begin{figure}[htbp]
  \vspace{-2ex}
  \centering
  \includegraphics[width=\textwidth]{./Figures/CCM_CarryChain.pdf}
  \vspace{-6ex}
  \caption{Four possible carry chain types in a constant coefficient multiplier with $n$-bit inputs. The notion $s$ denotes the shifted bits and $BP$ denotes the binary point.}
  \label{CCM_fig}
\end{figure}

@@

\subsection{Experimental Setup} \label{sub:experimental_setup}
We initially build up a test framework on an FPGA. The general architecture is depicted in Fig.~\ref{Test Framework}. The main body of the test framework consists of the circuit under test (CUT), the test frequency generator and the control logic, as shown in the dotted box in Fig.~\ref{Test Framework}. The I/Os of the CUT are registered by the launch registers (LRs) and the sample registers (SRs), which are all triggered by the test clock. Input test vectors are stored in the on-chip memory during initialization. The results are sampled using Xilinx ChipScope. Finally, we perform an offline comparison of the output of the original circuit at the rated frequency with the output of the overclocked as well as the truncated designs using the same input vectors.
%
\begin{figure}[htbp]
  \centering
  \vspace{-4ex}
  \includegraphics[width=3.5in]{./Figures/TestFramework.pdf}
  \vspace{-2ex}
  \caption{Test framework, which is composed of a measurement architecture (the dotted box) on an FPGA and an off-line comparator using software. The error-free data are obtained by either pre-computation or initial run with low frequencies.}
  \label{Test Framework}
  %\vspace{-1ex}
\end{figure}

@@

For example, the predicted frequencies and the actual frequencies of a $5^{th}$~order FIR filter using different word-lengths are depicted in Fig.~\ref{Fig:TimingMargin}. The ``actual'' maximum frequencies are computed by increasing the operating frequency from the rated value until errors are observed at the output; the maximum operating frequency with correct output is recorded for the current word-length. As can be seen in Fig.~\ref{Fig:TimingMargin}, the circuit can operate without errors at a much higher frequency in practice than predicted according to our experiments. A maximum speed differential of $3.2\times$ is obtained when the input signal is 5-bit.
%
\begin{figure}[ht]
  \centering
  \vspace{-2ex}
  \includegraphics[width=3.2in]{./Figures/Exp/FIR/RatedFrequency3.eps}
  %\vspace{-1.5ex}
  \caption{The maximum operating frequencies for different input word-lengths of an FIR filter. The dotted line depicts the rated frequency reported by the timing analysis tool. The solid line is obtained through real FPGA tests using our platform.}
  \label{Fig:TimingMargin}
  %\vspace{-3.5ex}
\end{figure}

@@

\section{Results and Discussion}\label{Section_Experiments}
\subsection{Case study: FIR filter}
We first assess the accuracy of our proposed models of errors. We record the value of MRE based on (\ref{Eq:MRE}) in a $5^{th}$~order FIR with respect to different operating frequencies. The modeled values of both overclocking error and truncation error of the FIR filter are presented in Fig.~\ref{Fig:FIR} (dotted lines), as well as the actual measurements on the FPGA (solid lines) with two types of input data. The results demonstrate that our models match well with the practical results obtained using the uniform independent inputs.
% FIR
\begin{figure}[tbp]
  \centering
  \vspace{-3ex}
  \includegraphics[width=3.5in]{./Figures/Exp/FIR/FIR_Error_New_TCAD.eps}
  \vspace{-2ex}
  \caption{A demonstration of two design perspectives with a $5^{th}$~order FIR filter, which is implemented on Virtex-6 FPGA. The modeled values of both overclocking errors and truncation errors are presented as dotted lines. The actual FPGA measurements are depicted using solid lines. Two types of inputs are employed in the overclocking scenario: the uniformly distributed data and the real image data from Lena.}
  \label{Fig:FIR}
\end{figure}

@@

In addition, we record the probability distribution of different length of carry chains using an 8-bit RCA with both data types, as shown in Fig.~\ref{Fig:CarryDistribution}. As the input data is 8-bit, the longest possible carry chain is of length 9-bit. It can be observed that when using Lena data,  the probability is only higher for carry chain with a length of 3-bit. While in other situations, using uniform data leads to higher probability, especially for long carry chain length. This finding explicitly demonstrates that longer carry chains happens with an even smaller probability when utilizing real data.
%
\begin{figure}[tbp]
  \centering
  \vspace{-2ex}
  \includegraphics[width=3.5in]{./Figures/CC_length2.eps}
  \vspace{-2ex}
  \caption{Probability distribution of different length of carry chains in a 8-bit RCA. For the uniform data, two inputs of RCA are randomly sampled from a uniform distribution. For the image data Lena, one input of RCA uses the original data and the other uses the delayed original data for several clock cycles.}
  \label{Fig:CarryDistribution}
\end{figure}
